---
title: "BumblebeeRobbing"
author: "Daniel"
date: "2025-06-09"
output: pdf_document
---

Behavioral assays done on bumblebees at Rocky Mountains Biology Laboratory (RMBL) field season 2023.
Observations were carried out by Nick Dabagia and Daniel Souto for Corydalis and Mertensia species.
Y-Tube assays carried out by Oriana Gutierrez in 2024.

Methods:

Refer to Souto-Vilarós et al. 'Yeast volatiles promote larceny in bumble bee behavior.'

Data analysis:

The dataset is arranged as date, plant species, bee individual, bee species, whether or not a choice was made, stalk number, treatment, whether or not the bee tried to legitimately visit the flower, whether it robbed the flower, time to rob, time feeding, flower number (total for that bout) and additional notes. 

Note that not all packages are used in the current analysis

```{r setup, include=FALSE, results=FALSE, warning=FALSE, message=TRUE}
rm(list=ls()) #removes all objects from work space - good practice to start all R sessions with this
library('ggplot2')
library('tidyverse')
library('dplyr')
library('stats')
library('multcomp')
library('lme4')
library('lmerTest')
library('gridExtra')
#library('ggpub')
visitation<-read.csv("data/bumblebee_behavior_data.csv", header = T)
```


We need to wrangle the data a bit. Namely:
Filter the data to include only cases where a flower was robbed (column CHOICE, "Yes"),
Remove an outlier which was inactive on a flwoer over 3 minutes. Make sure all data is numeric and there are no missing data.

This final data set includes: 57 total successful robber trials (out of 116 bee trials transcribed) 42 Corydalis 2° robbers (22 bifarius, 16 flavifrons, 4 mixtus) and 15 Mertensia 2° robbers (all flavifrons).

Bombus mixtus removed from final analysis due to low sample size.

```{r filtering data}
robbed_flowers <- visitation %>%
  filter(robbed == "yes") %>% #note that this will reduce total number of bees which made a choice since not all of them actually 'robbed'
  filter(species != "mixtus")%>%
  filter(timetorob !=154.7) 

robbed_flowers <- robbed_flowers %>%
  mutate(
    timetorob = as.numeric(timetorob),
    timefeeding = as.numeric(timefeeding)
  ) %>%
  filter(!is.na(timetorob) & !is.na(timefeeding))

robbed_flowers <- robbed_flowers %>%
  mutate(trytolegit_binary = ifelse(trytolegit == "yes", 1, 0))

#How many bees per bee species per flower species in the original and the filtered datasets?

visitation %>% group_by(sample, plantsp, species) %>% count()

robbed_flowers %>% group_by(sample, plantsp, species, bee) %>% count()
```

Simple summary statistics for time to rob (mean, SE) broken down by bumblebee species and plant sp. 
Note that time to rob is longer for control flowers, but also SE higher for controls. Consistent response for Mreu plants. Tactic switch, I just converted into a binary column yes = 1, no = 0. Feeding time is higher in both treatments, with high SE but consistent throughout.

```{r summary stats }
summary_stats <- robbed_flowers %>%
  group_by(species, treatment, plantsp) %>%
  summarize(
    Mean_TimeToRob = mean(timetorob, na.rm = TRUE),
    SE_TimeToRob = sd(timetorob, na.rm = TRUE) / sqrt(n()),
    Mean_TimeFeeding = mean(timefeeding, na.rm = TRUE),
    SE_TimeFeeding = sd(timetorob, na.rm = TRUE) / sqrt(n()),
    Mean_tacticswitch = mean(trytolegit_binary, na.rm = TRUE),
    SE_tacticswitch = sd(trytolegit_binary, na.rm = TRUE) / sqrt(n())
  ) %>% 
  arrange(plantsp, treatment)


print(summary_stats)

write.table(summary_stats, "results/summary_feeding_stats.csv", row.names = FALSE)
```


```{r}
# Summarize the data to calculate the number of visits
summary_data <- robbed_flowers %>%
  group_by(treatment, species, plantsp) %>%
  summarize(
    number_of_visits = n(),  # Count the number of visits
    .groups = "drop"
  )

# Calculate the mean, stdev, and se for each treatment group (control vs mreu)
visit_summary <- summary_data %>%
  group_by(species, plantsp) %>%
  summarize(
    mean_visits_control = mean(number_of_visits[treatment == "control"], na.rm = TRUE),
    mean_visits_treated = mean(number_of_visits[treatment == "mreu"], na.rm = TRUE),
    stdev_diff = sd(number_of_visits[treatment == "mreu"] - number_of_visits[treatment == "control"], na.rm = TRUE),
    se_diff = stdev_diff / sqrt(n()),
    .groups = "drop"
  )

# View the resulting summary table
print(visit_summary)
# Perform t-tests for each species and plantsp combination
t_test_results <- robbed_flowers %>%
  group_by(species, plantsp) %>%
  summarize(
    t_test = list(
      t.test(
        x = treatment == "mreu",    # Logical vector for mreu
        y = treatment == "control" # Logical vector for control
      )
    ),
    .groups = "drop"
  )

# Extract p-values and t-statistics from the t-test results
t_test_summary <- t_test_results %>%
  mutate(
    p_value = sapply(t_test, function(x) x$p.value),        # Extract p-value
    statistic = sapply(t_test, function(x) x$statistic)    # Extract t-statistic
  )

# Select only necessary columns
t_test_summary <- t_test_summary %>%
  dplyr::select(species, plantsp, p_value, statistic)

# View the summarized results
print(t_test_summary)

# Extract p-values, t-statistics, and degrees of freedom from the t-test results
t_test_summary <- t_test_results %>%
  mutate(
    p_value = sapply(t_test, function(x) x$p.value),        # Extract p-value
    statistic = sapply(t_test, function(x) x$statistic),    # Extract t-statistic
    df = sapply(t_test, function(x) x$parameter)            # Extract degrees of freedom
  )

# Explicitly use dplyr::select()
t_test_summary <- t_test_summary %>%
  dplyr::select(species, plantsp, p_value, statistic, df)

# View the summarized results
print(t_test_summary)
```

```{r}

# Create a histogram
ggplot(robbed_flowers, aes(x = interaction(species, plantsp), fill = treatment)) +
  geom_bar(position = "dodge", color = "black") +
  labs(x = NULL,
    y = "Number of Visits",
    fill = "Flower Treatment"
  ) +
  scale_fill_manual(
    values = c("mreu" = "#4169E1",
               "control" = "#FAFAD2")
  )+
  scale_y_continuous(expand = c(0, 0)) +  # Ensures bars rest on the x-axis line
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) -> plot

plot

```
```{r}

summary_df <- robbed_flowers %>%
  group_by(group = interaction(species, plantsp), treatment) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(group) %>%
  summarise(
    max_y = max(count),
    x_pos = as.numeric(factor(group)),  # for proper placement on x axis
    .groups = "drop"
  ) %>%
  mutate(label = "*")

ggplot(robbed_flowers, aes(x = interaction(species, plantsp), fill = treatment)) +
  geom_bar(position = position_dodge(width = 0.9), color = "black") +
  geom_text(
    data = summary_df,
    aes(x = group, y = max_y + 2, label = label),
    inherit.aes = FALSE,
    size = 6
  ) +
  labs(x = NULL,
       y = "Number of Visits",
       fill = "Flower Treatment") +
  scale_fill_manual(
    values = c("mreu" = "#4169E1", "control" = "#FAFAD2")
  ) +
  scale_y_continuous(expand = c(0, 0)) +  # Ensures bars rest on the x-axis line
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) -> plot2

plot2

ggsave("results/bee_visits_histogram.pdf", plot = plot2, width = 8, height = 6)

```


```{r time to rob boxplot}

mean_se <- function(x) {
  m <- mean(x, na.rm = TRUE)
  se <- sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
  return(paste0(round(m, 2), " ± ", round(se, 2)))
}

timetorob_box_species <- ggplot(robbed_flowers, aes(x = interaction(species, plantsp), y = timetorob, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
             aes(shape = treatment, color = treatment),
             size = 1, alpha = 0.75) +
  stat_summary(
    fun.data = function(x) {
      m <- mean(x)
      se <- sd(x) / sqrt(length(x))
      data.frame(y = m + se * -1.5, label = sprintf("%.2f ± %.2f", m, se))
    },
    geom = "text",
    position = position_dodge(width = 0.75),
    size = 3,
    vjust = 0,
    color = "black"
  ) +
  scale_fill_manual(
    values = c("mreu" = "#4169E1",
               "control" = "#FAFAD2")
  )+
  scale_color_manual(
    values = c("mreu" = "#4169E1",
               "control" = "#EEDD82")
  )+
  labs(
    title = "Time to rob by bee species and treatment",
    x = NULL,
    y = "Time to Rob"
  ) +
  #scale_fill_discrete(name = "Treatment") +
  #scale_color_discrete(name = "Treatment") +
  scale_shape_manual(name = "Treatment", values = c(16, 17), guide = "none") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 40)) +
  labs(
    title = "Time to rob by bee species and treatment",
    x = NULL,
    y = "Time to Rob"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black"),
    axis.line.x = element_line(),
    axis.line.y = element_line()
  )

timetorob <- timetorob_box_species +
  coord_cartesian(ylim = c(0, 40))

timetorob
ggsave("results/timetorob_pub.pdf", plot = timetorob, width = 8, height = 6)
```


ANOVA  time to rob on bee species (1) and plant species (2) - in both cases treatment is highly significant, but interaction between either plant or bee species and treatment is not. Suggests instead that only the nectar treatment significantly different

```{r ANOVA}
anova_result <- aov(timetorob ~ species + treatment, data = robbed_flowers)
summary(anova_result) 

anova_result_plant <- aov(timetorob ~ plantsp + treatment, data = robbed_flowers)
summary(anova_result_plant) 
```

To make sure this significance holds, I did wilcox test only on the mean time to rob. Again, significantly different (p = < 0.05)

```{r wilcox test}
treatment_control <- subset(robbed_flowers, treatment == "control") #control subset
treatment_mreu <- subset(robbed_flowers, treatment == "mreu") #inoculated flowers subset

# Perform the Wilcoxon rank-sum test
wilcox_test_result <- wilcox.test(treatment_control$timetorob, treatment_mreu$timetorob) #wilcox test

# Print the test result
print(wilcox_test_result) #wilcox result - highly significant difference of mean time to rob between treatments

```

We do the same analysis for feedint time and show that there is no significant difference between treatments on the feeding time of bees. It seems that yeast does not have a significant flavor(?) difference or not not different enough or nasty enough for bees to care. They spend the same amount of time feeding on sterile or yeast innoculated nectar

```{r feeding time}
feedingtime_box_species <- ggplot(robbed_flowers, aes(x = interaction(species, plantsp), y = timefeeding, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) + 
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
             aes(shape = treatment, color = treatment),
             size = 1, alpha = 0.75) +
  stat_summary(
    fun.data = function(x) {
      m <- mean(x)
      se <- sd(x) / sqrt(length(x))
      data.frame(y = m + se * -.5, label = sprintf("%.2f ± %.2f", m, se))
    },
    geom = "text",
    position = position_dodge(width = 0.75),
    size = 3,
    vjust = 0,
    color = "black"
  ) +
  scale_fill_manual(
    values = c("mreu" = "#4169E1",
               "control" = "#FAFAD2")
  )+
  scale_color_manual(
    values = c("mreu" = "#4169E1",
               "control" = "#EEDD82")
  )+
  labs(
    title = "Feeding time by bee species and treatment",
    x = NULL,
    y = "Time to Rob"
  ) +
  #scale_fill_discrete(name = "Treatment") +
  #scale_color_discrete(name = "Treatment") +
  scale_shape_manual(name = "Treatment", values = c(16, 17), guide = "none") +
  scale_y_continuous(expand = c(0, 0)) +  # Ensures bars rest on the x-axis line
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black"),
    axis.line.x = element_line(),
    axis.line.y = element_line(),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

feedingtime_box_species

ggsave("results/feeding_pub.pdf", plot = feedingtime_box_species, width = 8, height = 6)

```

ANOVA and wilcox test confirm. No difference on time spent feeding depending on treatment of nectar.

```{r ANOVA and Wilcox feeding time}
anova_result_feeding <- aov(timefeeding ~ species + treatment, data = robbed_flowers)

# Print the ANOVA table
summary(anova_result_feeding)

# Perform the Wilcoxon rank-sum test
wilcox_test_result_feeding <- wilcox.test(treatment_control$timefeeding, treatment_mreu$timefeeding)

# Print the test result
print(wilcox_test_result_feeding)
```

This included all plant species combined.
Below I break down everything per plant species (Corydalis and Mertensia)

First subset data 

```{r subset per plants}
corydalis <- subset(robbed_flowers, plantsp == "corydalis")
mertensia <- subset(robbed_flowers, plantsp == "mertensia")
```

Since for Corydalis we have 2 bee species, I still did ANOVA. For Mertensia, there is only one bee species so makes more sense to do wilcox. In the case of Corydalis, the two bee species take different times to find the robbing hole (p = 0.03), while time to find the hole for different treatments is stil highly significant (p =0.001). We could do different analysis for each bee species, if worth it.

```{r Corydalis ANOVA}
anova_result_robbing_corydalis <- aov(timetorob ~ species + treatment, data = corydalis)
# Print the ANOVA table
summary(anova_result_robbing_corydalis)

anova_result_feeding_corydalis <- aov(timefeeding ~ species + treatment, data = corydalis)
# Print the ANOVA table
summary(anova_result_feeding_corydalis)
```
For Corydalis, significant difference both for species and treatment, but not the interaction. 
So, bee species show difference in the time it takes to find a robbing hole (p = 0.03), and significantly different according to treatment (p = 0.001).

Wilcox test for mertenisa, time to rob significantly different between treatments (P = 0.006)
```{r wilcox for mertensia}

wilcox_mertensia_timetorob <- wilcox.test(timetorob ~ treatment, data = mertensia)
wilcox_mertensia_timetorob

```

```{r }
legit_summary <- corydalis %>%
  group_by(unique_name) %>%
  summarise(trytolegit_binary = sum(trytolegit_binary))

print(legit_summary)

visit_summary <- corydalis %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))

visit_summary
```

```{r}
y_tube <- read.csv("data/y_tube_ori_2025.csv", header = T)

table(y_tube$Initial_Choice)
chisq.test(table(y_tube$Initial_Choice))
 
t <- t.test(y_tube$Yeast_time, y_tube$Control_time, paired = TRUE, alternative = "greater") 
t
```

```{r}

corydalis_bifarius <- subset(corydalis, species == "bifarius")
corydalis_flavifrons <- subset(corydalis, species == "flavifrons")

summary_corydalis <- corydalis %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))


# Create a new column indicating whether trytolegit_binary > 0
legit_summary <- legit_summary %>%
  mutate(trytolegit_gt_0 = ifelse(trytolegit_binary > 0, "Trytolegit > 0", "Trytolegit = 0"))

# Summarize the data by the new column
summary_data <- legit_summary %>%
  group_by(trytolegit_gt_0) %>%
  summarise(total_visits = n())

# Calculate total visits
total_visits <- sum(summary_data$total_visits)

# Calculate percentage of visits for each category
summary_data <- summary_data %>%
  mutate(percentage = total_visits / sum(total_visits))

# Plot the barplot
ggplot(summary_data, aes(x = "", y = total_visits, fill = trytolegit_gt_0)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = scales::percent(percentage)), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 5) +
  labs(x = NULL, y = "Total Visits") +
  scale_fill_manual(values = c("Trytolegit = 0" = "gray", "Trytolegit > 0" = "black")) +
  theme_minimal() +
  theme(axis.text.x=element_blank())  # Hide x-axis label


```

```{r }
legit_summary_mert <- mertensia %>%
  group_by(unique_name) %>%
  summarise(trytolegit_binary = sum(trytolegit_binary))

print(legit_summary_mert)

visit_summary_mert <- mertensia %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))

visit_summary_mert
```
 
```{r}
summary_mertensia <- mertensia %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))


# Create a new column indicating whether trytolegit_binary > 0
legit_summary_mert <- legit_summary_mert %>%
  mutate(trytolegit_gt_0 = ifelse(trytolegit_binary > 0, "Trytolegit > 0", "Trytolegit = 0"))

# Summarize the data by the new column
summary_data_mert <- legit_summary_mert %>%
  group_by(trytolegit_gt_0) %>%
  summarise(total_visits = n())

# Calculate total visits
total_visits_mert <- sum(summary_data_mert$total_visits)

# Calculate percentage of visits for each category
summary_data_mert <- summary_data_mert %>%
  mutate(percentage = total_visits / sum(total_visits))

# Plot the barplot
ggplot(summary_data_mert, aes(x = "", y = total_visits, fill = trytolegit_gt_0)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = scales::percent(percentage)), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 5) +
  labs(x = NULL, y = "Total Visits") +
  scale_fill_manual(values = c("Trytolegit = 0" = "gray", "Trytolegit > 0" = "black")) +
  theme_minimal() +
  theme(axis.text.x=element_blank())  # Hide x-axis label
```

```{r}
summary_cory_bif <- corydalis_bifarius %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))
legit_summary_corybif <- corydalis_bifarius %>%
  group_by(unique_name) %>%
  summarise(trytolegit_binary = sum(trytolegit_binary))

visit_summary_corybif <- corydalis_bifarius %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))


# Create a new column indicating whether trytolegit_binary > 0
legit_summary_corybif <- legit_summary_corybif %>%
  mutate(trytolegit_gt_0 = ifelse(trytolegit_binary > 0, "Trytolegit > 0", "Trytolegit = 0"))

# Summarize the data by the new column
summary_data_corybif <- legit_summary_corybif %>%
  group_by(trytolegit_gt_0) %>%
  summarise(total_visits = n())

# Calculate total visits
total_visits_corybif <- sum(summary_data_corybif$total_visits)

# Calculate percentage of visits for each category
summary_data_corybif <- summary_data_corybif %>%
  mutate(percentage = total_visits / sum(total_visits))

# Plot the barplot
ggplot(summary_data_corybif, aes(x = "", y = total_visits, fill = trytolegit_gt_0)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = scales::percent(percentage)), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 5) +
  labs(x = NULL, y = "Total Visits B. bifarius on Corydalis") +
  scale_fill_manual(values = c("Trytolegit = 0" = "gray", "Trytolegit > 0" = "black")) +
  theme_minimal() +
  theme(axis.text.x=element_blank())  # Hide x-axis label

```

```{r}
 summary_cory_flav <- corydalis_flavifrons %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))
legit_summary_coryflav <- corydalis_flavifrons %>%
  group_by(unique_name) %>%
  summarise(trytolegit_binary = sum(trytolegit_binary))

visit_summary_coryflav <- corydalis_flavifrons %>%
  group_by(unique_name) %>%
  summarise(total_visits = n(),
            legit_visits = sum(trytolegit == 'yes'))


# Create a new column indicating whether trytolegit_binary > 0
legit_summary_coryflav <- legit_summary_coryflav %>%
  mutate(trytolegit_gt_0 = ifelse(trytolegit_binary > 0, "Trytolegit > 0", "Trytolegit = 0"))

# Summarize the data by the new column
summary_data_coryflav <- legit_summary_coryflav %>%
  group_by(trytolegit_gt_0) %>%
  summarise(total_visits = n())

# Calculate total visits
total_visits_coryflav <- sum(summary_data_coryflav$total_visits)

# Calculate percentage of visits for each category
summary_data_coryflav <- summary_data_coryflav %>%
  mutate(percentage = total_visits / sum(total_visits))

# Plot the barplot
ggplot(summary_data_coryflav, aes(x = "", y = total_visits, fill = trytolegit_gt_0)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = scales::percent(percentage)), 
            position = position_stack(vjust = 0.5), 
            color = "white", size = 5) +
  labs(x = NULL, y = "Total Visits B. flavifrons on Corydalis") +
  scale_fill_manual(values = c("Trytolegit = 0" = "gray", "Trytolegit > 0" = "black")) +
  theme_minimal() +
  theme(axis.text.x=element_blank())  # Hide x-axis label
```
 
